---
title: "In-class Exercise 4: Interactive K-function"
author: "Deepika Raja"
date: "September 20, 2025"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# **1. Overview**

**Geographically Weighted Summary Statistics (GWSS)** compute local measures (e.g., mean, median, standard deviation) instead of global ones, by assigning higher weights to nearby data points using a kernel and bandwidth. This highlights **spatial heterogeneity**, showing how patterns and relationships vary across different locations.

# **2. Application of GWSS:**

GWSS are used to capture how variable relationships differ across space in domains like environment, health, and socio-economics. They help reveal localized patterns—such as ecological changes, health disparities, or human impacts on landscapes—by building local rather than global models. This exposes spatial non-stationarity and supports more precise resource management and policy decisions.

# **3.Loading the package**

To complete this exercise, sf, spdep, tmap, tidyverse, knitr and GWmodel will be used.

```{r}
pacman::p_load(sf, ggstatsplot, tmap, tidyverse, knitr, GWmodel)
```

# **4. Preparing the Data**

The following datasets will be used in this exercise:

| **Data Set** | **Description** | **Format** |
|----|----|----|
| **Hunan County Boundary Layer** | A geospatial dataset containing Hunan’s county boundaries. | ESRI Shapefile |
| **Hunan_2012.csv** | A CSV file containing selected local development indicators for Hunan in 2012. | CSV |

```{r}
hunan_sf <- st_read(dsn = "data/geospatial",
                 layer = "Hunan")
```

```{r}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")
```

```{r}
glimpse(hunan2012)
```

\<!--# Note:

Recall that to do left join, we need a common identifier between the 2 data objects. The content must be the same eg. same format and same case. In Hands-on Exercise 1B, we need to (PA, SZ) in the dataset to uppercase before we can join the data. --\>

```{r}
hunan_sf <- left_join(hunan_sf, hunan2012) %>%
  select(1:3, 7, 15, 16, 31, 32)
```

```{r}
hunan_sf
```

# **5 Mapping GDPPC**

To plot a chrolopleth map of geographic distribution of GDP per Capita (GDPPC) in Hunan:

```{r}
basemap <- tm_shape(hunan_sf) +
  tm_polygons() +
  tm_text("NAME_3", size=0.5)

gdppc <- qtm(hunan_sf, "GDPPC")
tmap_arrange(basemap, gdppc, asp=1, ncol=2)
```

# **6. Converting to SpatialPolygonDataFrame**

\<!--# **Note**

GWmodel presently is built around the older sp and not sf formats for handling spatial data in R --\>

```{r}
hunan_sp <- hunan_sf %>%
  as_Spatial()
```

# **7 Geographically Weighted Summary Statistics with Adaptive Bandwidths**

In this section, we aim to determine the optimal adaptive bandwidth for performing Geographically Weighted Regression (GWR). Specifically, we are interested in finding the best bandwidth to use for summarizing the spatial variation in GDP per capita (GDPPC) across the Hunan region.

## **7.a Determine Adaptive Bandwidth**

An adaptive bandwidth allows the number of neighbors considered in the model to vary depending on the density of data points. This is particularly useful when data points are unevenly distributed across the study area.

We will use two different criteria—cross-validation (CV) and [Akaike information criterion](https://en.wikipedia.org/wiki/Akaike_information_criterion) to determine the optimal bandwidth.

**The bandwidth that minimizes these metrics will be selected.**

### **7.a.1. Cross Validation**

```{r}
bw_CV<- bw.gwr(GDPPC ~ 1,
               data = hunan_sp,
               approach= "CV",
               adaptive = TRUE,
               kernel = "bisquare",
               longlat = T)  # great circle distance
```

```{r}
bw_CV
```

### **7.a.2. Akaike Information Criterion (AIC)**

Next, we use the AIC approach to determine the optimal bandwidth. AIC is a model selection criterion that balances model fit and complexity, with a lower AIC value indicating a better model.

We use the same GWR model setup, but the bandwidth is now optimized based on the AIC value instead of cross-validation.

```{r}
bw_AIC<- bw.gwr(GDPPC ~ 1,
               data = hunan_sp,
               approach= "AIC",
               adaptive = TRUE,
               kernel = "bisquare",
               longlat = T)
```

```{r}
bw_AIC
```

**\
**\<!--# **Note**

**Intepretation**

The output from these 2 methods indicate the number of nearest neighbour we should choose. In this case, both methods produce the same result: 22 nearest neighbours.

Sometimes the result may differ, and either methods is acceptable for further analysis. --\>

# **8 Geographically Weighted Summary Statistics with adaptive bandwidth**

To compute Geographically Weighted Summary Statistics:

```{r}
gwstat <- gwss (data = hunan_sp,
                vars = "GDPPC",
                bw = bw_AIC,
                kernel = "bisquare",
                adaptive = TRUE,
                longlat = T)
```

\<!--# **Note**

We use `bw_AIC` as the bandwidth parameter, which was determined previously based on AIC optimization.

Additionally, we apply the same `bisquare` kernel for consistency with the CV and AIC computation above.

The output of the `gwss()` function is a `gwss object`, which is a **list containing localized summary statistics for GDPPC across Hunan**.

Note that the abbreviation in the output refers to:

-   LM : local mean

-   LSD: local standard deviation

-   LVar: local variance

-   LSKe: standard estimations

-   LCV: local correlation variance --\>

## **8.a. Preparing the output data**

Let’s observe the `gwstat` object before converting to a suitable format for analysis.

```{r}
class(gwstat)
```

```{r}
gwstat
```

```{r}
gwstat$SDF
```

In particular, we are interested to extract the `SDF` data table from `gwstat`. We can convert it into a data frame and append it onto `hunan_sf`.

```{r}
gwstat_df <- as.data.frame(gwstat$SDF)
hunan_gstat <- cbind(hunan_sf, gwstat_df)
```

## **8.b. Visualising geographically weighted summary statistics**

```{r}
tm_shape(hunan_gstat) +
  tm_polygons(fill = "GDPPC_LM",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "local Mean")) +
  tm_borders(fill_alpha = 0.5) +
  tm_layout(main.title = "Distribution of geographically wieghted mean",
            main.title.position = "center",
            main.title.size = 2.0,
            frame = TRUE)
```

# **9 Geographically Weighted Summary Statistics with Fixed Bandwidth**

Similarly, we can use the same process to generate summary stats with fixed bandwidth.

## **9.a. Determine Fixed Bandwidth**

1.  Cross-Validation

```{r}
bw_CV_fixed <- bw.gwr(GDPPC ~ 1,
             data = hunan_sp,
             approach = "CV",
             adaptive = FALSE,
             kernel = "bisquare",
             longlat = T)
```

```{r}
bw_CV_fixed
```

2.  AIC

```{r}
bw_AIC_fixed <- bw.gwr(GDPPC ~ 1,
             data = hunan_sp,
             approach ="AIC",
             adaptive = FALSE,
             kernel = "bisquare",
             longlat = T)
```

```{r}
bw_AIC_fixed
```

\<!--# **\
Note**

Note the results differs this time.

We will just use `bw_AIC_fixed` for this example.

--\>

```{r}
gwstat_fixed <- gwss(data = hunan_sp,
               vars = "GDPPC",
               bw = bw_AIC_fixed,
               kernel = "bisquare",
               adaptive = FALSE,
               longlat = T)
```

## **9.b. Preparing the output data**

```{r}
gwstat_df_fixed <- as.data.frame(gwstat_fixed$SDF)
hunan_gstat_fixed <- cbind(hunan_sf, gwstat_df_fixed)
```

## **9.c. Visualising geographically weighted summary statistics**

```{r}
tm_shape(hunan_gstat) +
  tm_polygons(fill = "GDPPC_LM",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "local Mean")) +
  tm_borders(fill_alpha = 0.5) +
  tm_layout(main.title = "Distribution of geographically wieghted mean",
            main.title.position = "center",
            main.title.size = 2.0,
            frame = TRUE)
```

# **10 Visualizing Correlation**

**Business question: Is there any relationship between GDP per capita and Gross Industry Output?**

```{r}
ggscatterstats(
  data = hunan2012,
  x = Agri,
  y = GDPPC,
  xlab = "Gross Agriculture Output",
  ylab = "GDP per capita",
  label.var = County,
  label.expression = Agri > 10000 & GDPPC > 50000,
  point.label.args = list(alpha = 0.7, size = 4, color = "grey50"),
  xfill = "#CC79A7",
  yfill = "#009E73",
  title = "Relationship between GDP PC and Gross Agriculture Output")
```

\<!--# **\
Note**

Note that above shows a **conventional statistical solution** to the business question. We can also approach the same question with a geospatial approach.

--\>

# **11 Geographically Weighted Correlation with Adaptive Bandwidth**

To come up with the geospatial analytics solution, we can repeat what we have learnt above.

```{r}
# determine bandwidth
bw <- bw.gwr(GDPPC ~ GIO,
             data = hunan_sp,
             approach = "AICc",
             adaptive = TRUE)
```

```{r}
# compute gwCorrelation
gwstats <- gwss(hunan_sp,
                vars = c("GDPPC", "GIO"),
                bw = bw,
                kernel = "bisquare",
                adaptive = TRUE,
                longlat = T)

gwstats$SDF
```

```{r}
# convert result to df
gwstat_df <- as.data.frame(gwstats$SDF) %>%
  # select(c(12,13)) %>%
  select(c("Corr_GDPPC.GIO","Spearman_rho_GDPPC.GIO")) %>%
  rename(gwCorr = Corr_GDPPC.GIO,
         gwSpearman = Spearman_rho_GDPPC.GIO)

hunan_Corr <- cbind(hunan_sf, gwstat_df)
hunan_Corr
```

## **11.a. Visualizing Local Correlation**

1.  Local Correlation Coefficient

```{r}
tm_shape(hunan_Corr) +
  tm_polygons(fill = "gwCorr",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "local Correlation Coefficient")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Local Correlation Coefficient", 
  size = 2.0) +
  tm_layout(frame = TRUE)
```

\<!--# **\
Note**

**Interpretation**

-   The strongest correlations are found in the eastern and northern parts of the province, indicated by the darker shades.

-   The weaker correlations are located in the central and western areas, where the lighter colors predominate.

--\>

2.  Local Spearman Coefficient

Note that we will observe similar trend using Local Spearman Coefficient. See notes below.

```{r}
tm_shape(hunan_Corr) +
  tm_polygons(fill = "gwSpearman",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "local Spearman Rho")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Local Spearman Rho", 
  size = 2.0) +
  tm_layout(frame = TRUE)
```

\<!--# **\
Tip**

**Notes on local correlation coefficient and the local Spearman coefficient**:

1.  Local Correlation Coefficient (Pearson)

-   **Type**: Parametric

    -   The local correlation coefficient, often represented by Pearson’s correlation coefficient, assumes that the data follows a normal distribution.

-   **Nature**: Continuous

    -   It measures the linear relationship between two continuous variables.

-   **Type of Measure**: Not Ranked

    -   The Pearson correlation is sensitive to the actual values of the data points, not their ranks. It considers both the magnitude and direction of the linear relationship.

2.  Local Spearman Coefficient

-   **Type**: Non-Parametric

    -   The local Spearman coefficient is a rank-based measure and does not assume any specific distribution for the data. It is robust to non-normality.

-   **Nature**: Continuous (based on ranks)

    -   Although it works with ranks, the coefficient itself can take any continuous value between -1 and 1, like Pearson’s.

-   **Type of Measure**: Ranked

    -   The Spearman correlation is based on the ranks of the data rather than their actual values. It measures the strength and direction of a monotonic relationship between two variables.

--\>

# **12 Reference**

Kam, T. S. Geographically Weighted Summary Statistics - gwModel methods. *ISSS626 Geospatial Analytics and Applications.* <https://isss626-ay2025-26aug.netlify.app/in-class_ex/in-class_ex04/in-class_ex04#/title-slide>
