---
title: "In-class Exercise 6: Emerging Hot Spot Analysis"
author: "Deepika Raja"
date: "October 4, 2025"
date-modified: "Ocr 5, 2025"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# 1.Overview

Emerging Hot Spot Analysis (EHSA) is a spatio-temporal analysis method for revealing and describing how hot spot and cold spot areas evolve over time. The analysis consist of four main steps:

Building a space-time cube, Calculating Getis-Ord local Gi\* statistic for each bin by using an FDR correction, Evaluating these hot and cold spot trends by using Mann-Kendall trend test, Categorising each study area location by referring to the resultant trend z-score and p-value for each location with data, and with the hot spot z-score and p-value for each bin.

# **2. Getting started**

## **Installing and Loading the R Packages**

As usual, `p_load()` of **pacman** package will be used to check if the necessary packages have been installed in R, if yes, load the packages on R environment.

Five R packages are need for this in-class exercise, they are: sf, sfdep, tmap, plotly, and tidyverse.

```{r}
pacman::p_load(sf, sfdep, tmap, 
               plotly, tidyverse, 
               Kendall)
```

# **3. The Data**

For the purpose of this in-class exercise, the Hunan data sets will be used. There are two data sets in this use case, they are:

-   Hunan, a geospatial data set in ESRI shapefile format, and

-   Hunan_GDPPC, an attribute data set in csv format.

Before getting started, reveal the content of *Hunan_GDPPC.csv* by using Notepad and MS Excel.

### **3.a. Importing geospatial data**

In the code chunk below, `st_read()` of **sf** package is used to import *Hunan* shapefile into R.

```{r}
hunan <- st_read(dsn = "data/geospatial", 
                 layer = "Hunan")
```

### **Importing attribute table**

In the code chunk below, `read_csv()` of **readr** is used to import *Hunan_GDPPC.csv* into R.
