---
title: "In class Exercise 9 : Modelling Geographic of Accessibility"
author: "Deepika Raja"
date: "November 08, 2025"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  message: false
  freeze: true
  warning: false
format: 
  html:
    code-fold: false
    code-summary: "code chunk"
    number-sections: true
    code-overflow: scroll
    number-depth: 4
    code-line-numbers: false
---

## 1. Introduction

We're going to model geographic accessibility, which is basically how easy it is for people to get to a specific place or service. We'll measure this using travel time and distance in R.

### 1.a. Learning Outcome:

-   Import GIS polygon data into R and save them as *simple feature data frame* by using appropriate functions of **sf** package of R;
-   Import aspatial data into R and save them as *simple feature data frame* by using appropriate functions of **sf** package of R;
-   Computer accessibility measure by using **Hansen’s potential model** and **Spatial Accessibility Measure** (SAM); and
-   Visualise the accessibility measures by using **tmap** and **ggplot2** packages.

### 1.b. The Data

+--------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+----------------+
| Dataset Name             | Description                                                                                                                                                                                                                                                                                                         | Format         |
+==========================+=====================================================================================================================================================================================================================================================================================================================+================+
| `MP14_SUBZONE_NO_SEA_PL` | URA Master Plan 2014 subzone boundary GIS data, downloaded from [data.gov.sg](data.gov.sg).                                                                                                                                                                                                                         | ERSI Shapefile |
+--------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+----------------+
| `hexagons`               | A 250m radius hexagons GIS data, created by using [`st_make_grid()`](https://r-spatial.github.io/sf/reference/st_make_grid.html) of **sf** package.                                                                                                                                                                 | ESRI shapefile |
+--------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+----------------+
| `ELDERCARE`              | Locations of eldercare service, downloaded from [data.gov.sg](https://data.gov.sg/collections/714/view). Two formats were available, ESRI shapefile and Google KML. The Shapefile format was selected as it is a geospatial-native format that preserves coordinate systems and spatial attributes more accurately. | ESRI shapefile |
+--------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+----------------+
| `OD_Matrix`              | A travel distance matrix between the origin hexagon cells and eldercare service locations, measured by **metres**. It includes:                                                                                                                                                                                     | CSV            |
|                          |                                                                                                                                                                                                                                                                                                                     |                |
|                          | -   `origin_id`: hexagon identifier                                                                                                                                                                                                                                                                                 |                |
|                          |                                                                                                                                                                                                                                                                                                                     |                |
|                          | -   `destination_id`: eldercare facility identifier                                                                                                                                                                                                                                                                 |                |
|                          |                                                                                                                                                                                                                                                                                                                     |                |
|                          | -   `entry_cost`: perpendicular distance from the origin to nearest road entry                                                                                                                                                                                                                                      |                |
|                          |                                                                                                                                                                                                                                                                                                                     |                |
|                          | -   `network_cost`: actual travel distance on the road                                                                                                                                                                                                                                                              |                |
|                          |                                                                                                                                                                                                                                                                                                                     |                |
|                          | -   `exit_cost`: perpendicular distance from destination to nearest road connection                                                                                                                                                                                                                                 |                |
|                          |                                                                                                                                                                                                                                                                                                                     |                |
|                          | -   `total_cost`: total travel cost, computed as `entry_cost+network_cost+exit_cost`                                                                                                                                                                                                                                |                |
+--------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+----------------+

## 2. Getting Started

There are **six** packages to be used in this exercise.

+---------------------------------------------------------------------------------------------------------------------------------------------------------+----------------------------------------------------------------------------------------------------------------------------+
| Package                                                                                                                                                 | Description                                                                                                                |
+=========================================================================================================================================================+============================================================================================================================+
| [sf](https://r-spatial.github.io/sf/)                                                                                                                   | Handles importing, managing, and processing vector-based geospatial data                                                   |
+---------------------------------------------------------------------------------------------------------------------------------------------------------+----------------------------------------------------------------------------------------------------------------------------+
| [tidyverse](https://www.tidyverse.org/)                                                                                                                 | A collection of R packages for data wrangling and visualisation, including **readr**, **dplyr**, and **ggplot2**.          |
+---------------------------------------------------------------------------------------------------------------------------------------------------------+----------------------------------------------------------------------------------------------------------------------------+
| [tmap](https://r-tmap.github.io/tmap/)                                                                                                                  | Creates cartographic quality static and interactive thematic (choropleth) maps.                                            |
+---------------------------------------------------------------------------------------------------------------------------------------------------------+----------------------------------------------------------------------------------------------------------------------------+
| [Spatial](https://cran.r-project.org/web/packages/SpatialML/index.html)[ACC](https://cran.r-project.org/web/packages/SpatialAcc/refman/SpatialAcc.html) | Provides a set of spatial accessibility measures from demand to supply                                                     |
+---------------------------------------------------------------------------------------------------------------------------------------------------------+----------------------------------------------------------------------------------------------------------------------------+
| [ggstatsplot](https://indrajeetpatil.github.io/ggstatsplot/)                                                                                            | An extension of [`{ggplot2}`](https://github.com/tidyverse/ggplot2) package to create enriched plots for statistical tests |
+---------------------------------------------------------------------------------------------------------------------------------------------------------+----------------------------------------------------------------------------------------------------------------------------+
| [reshape2](https://www.rdocumentation.org/packages/reshape2/versions/1.4.4)                                                                             | Provides tools to restructure and combine data.                                                                            |
+---------------------------------------------------------------------------------------------------------------------------------------------------------+----------------------------------------------------------------------------------------------------------------------------+

Let's install and launch these packages into R environment.

```{r}
pacman::p_load(tmap, SpatialAcc, sf, ggstatsplot, reshape2, tidyverse)
```

## 3. Geospatial Data Wrangling

### 3.a. Importing Geospatial Data

The code chunk below uses `st_read()` of **sf** package to import the three ERSI Shapefile files into R as `sf` objects.

```{r}
mpsz <- st_read(dsn = "data/geospatial", layer = "MP14_SUBZONE_NO_SEA_PL")
```

```{r}
hexagons <- st_read(dsn = "data/geospatial", layer = "hexagons") 
```

```{r}
eldercare <- st_read(dsn = "data/geospatial", layer = "ELDERCARE") 
```

::: insightbox
The import reports reveal the geometry type of each file, and improtant information like dimension and projected CRS. We notice `mpsz` file does not have proper "SVY21/ Singapore TM".
:::

### 3.b. Updating CRS Information

The code chunk below assigns the correct ESPG(3414) to our newly imported `sf` objects.

```{r}
mpsz <- st_transform(mpsz, 3414)
eldercare <- st_transform(eldercare, 3414)
hexagons <- st_transform(hexagons, 3414)
```

Then we can verify the update via calling `st_crs()` of **sf** package. The EPSG has been updated accordingly.

```{r}
st_crs(mpsz)
```

We shall quickly plot map for eldercare centers in Singapore.

```{r}
tm_shape(mpsz) +
  tm_polygons(alpha = 0.2, border.col = "gray") +
  tm_dots(col = "black", size = 0.1, alpha = 0.6) +
  tm_title("Eldercare Facilities Distribution")
```

### 3.c. Cleaning and Updating Attribute fields of the Geospatial Data

After inspecting the origin and destination datasets, we reorganise the attributes and retain only the required ID fields. We then create a **demand** field for the hexagon layer, representing the estimated number of people who require eldercare services within each hexagonal area. Likewise, we add a **capacity** field for the eldercare layer, which indicates the service capacity or supply at each facility (e.g., number of beds or service slots available).

For the purpose of this exercise, both demand and capacity are assigned a constant value of **100** to simplify the demonstration. In practice, actual population data and real facility capacity figures should be used.

```{r}
eldercare <- eldercare %>%
  select(fid, ADDRESSPOS) %>%
  mutate(capacity = 100)
```

```{r}
hexagons <- hexagons %>%
  select(fid) %>%
  mutate(demand = 100)
```

## 4. Apsatial Data Handling and Wrangling

### 4.a. Importing Distance Matrix

The code chunk below uses `read_cvs()` of **readr** package to import the matrix dataset into R as a tibble data.frame .

```{r}
ODMatrix <- read_csv("data/aspatial/OD_Matrix.csv", skip = 0)
```

### 4.b. Tidying Distance Matrix

The imported ODMatrix is structured that the cost component appears as a separate column.

```{r}
knitr::kable(head(ODMatrix, 10))
```

A standardised matrix format stores origins as rows and destinations as columns, allowing the distance between any origin–destination pair to be retrieved by referencing the corresponding row and column indices.

This transformation can be performed using the `spread()` or `pivot_wider()` function from the **tidyr** package to convert the long (thin) table format into a wide (fat) matrix format.

```{r}
distmat <- ODMatrix %>%
  select(origin_id, destination_id, total_cost) %>%
  pivot_wider(names_from = destination_id, values_from = total_cost) %>% #spread(destination, total_cost)
  select(-'origin_id')  #remove the column origin_id
```

Next, we rescale the distance values from metres to kilometres. `as.matrix()` is used here to convert the `tbl_df` data frame into a numeric `matrix` format.

```{r}
distmat_km <- as.matrix(distmat/1000)
```

## 5. Hansen Method

ac() computes spatial accessibility of demand locations (e.g., residential areas) to service locations (e.g., eldercare centres). It implements several well-known measures:

-   "Hansen":Hansen gravity-type accessibility

-   "KD2SFCA": Kernel/Downgraded 2SFCA (distance-decay)

-   "SAM": Spatial Accessibility Measure

**Usage**:

-   ac(p, n, D, d0, power=2, family="SAM")

**Arguments**:

-   p: numeric vector of demand at each origin (e.g., number of senior residents per hexagon).

-   n: numeric vector of supply at each destination (e.g., number of beds per eldercare centre).

-   D: distance or time matrix separating origins (rows) from destinations (columns). Prefer road-network distance or travel time.

-   d0: catchment threshold (distance/time) defining which OD pairs “count”.

-   power: distance-decay exponent (commonly 2 in gravity models).

-   family: which accessibility method to compute: "SAM", "2SFCA", "KD2SFCA", or "Hansen" (default "SAM").

All four methods are **gravity-based** accessibility measures where accessibility **improves** with greater service capacity and proximity, and **decreases** with distance and population demand due to competition for limited resources.

**Return value**

A numeric vector of accessibility scores for each origin (length = length(p)).

## 6. Modelling and Visualising Accessibility using Hansen Method

The code chunk below uses `ac()` to compute the Hansen’s accessibility and saves the output in a data frame by using data.frame().

```{r}
acc_Hansen <- data.frame(ac(hexagons$demand,
                            eldercare$capacity,
                            distmat_km,
                            power = 0.5,
                            family = "Hansen"))
```

```{r}

# Create a list to store all results
hansen_results <- list()

# Calculate accessibility for each power value
power_values <- c(0.5, 1, 1.5, 2, 2.5)

for (power_val in power_values) {
  acc_temp <- data.frame(ac(hexagons$demand,
                            eldercare$capacity,
                            distmat_km,
                            power = power_val,
                            family = "Hansen"))
  colnames(acc_temp) <- paste0("accHansen_power", power_val)
  hansen_results[[as.character(power_val)]] <- acc_temp
}

# Combine all results
acc_Hansen_all <- bind_cols(hansen_results)

# For your existing workflow, keep one version (e.g., power = 2)
acc_Hansen <- data.frame(ac(hexagons$demand,
                            eldercare$capacity,
                            distmat_km,
                            power = 2,
                            family = "Hansen"))
```

Let's have a quick review on the Hansen accessibility measure, after rename the column name to `accHansen`.

```{r}
colnames(acc_Hansen) <- "accHansen" 
```

```{r}
acc_Hansen <- acc_Hansen %>%
  as_tibble()
```

```{r}
knitr::kable(head(acc_Hansen, 10), 
             align = 'c',  # column alignment: c=center, l=left, r=right
             format = "html")  # or "latex", "markdown", "pandoc"
```

Lastly, we join the `acc_Hansen` `tibble` with the `hexagon` `sf` data frame using `bind_cols()`. Because the `sf` object is on the **left** side of the operation, the resulting dataset automatically inherits the **sf** class, preserving its spatial attributes.

```{r}
hexagon_Hansen <- bind_cols(hexagons, acc_Hansen)
```

```{r}
knitr::kable(head(hexagon_Hansen,10),
             align = "c",
            format = "markdown")
```

### 6.a. Visualising Hansen’s Accessibility

#### 6.a.1. Extracting Map Extend

First, we extract the extent of the hexagon sf layer using `st_bbox()` of sf package.

```{r}
mapex <- st_bbox(hexagons)
```

Next, we create a choropleth map with the **tmap** package to visualise eldercare service accessibility across Singapore.

```{r}
tm_shape(hexagon_Hansen,
         bbox = mapex) + 
  tm_polygons(fill = "accHansen", 
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 10,
                values = "brewer.blues")) +
tm_shape(eldercare) +
  tm_dots(size = 0.3,
          fill = "red",
          col = "black") +
  tm_title("Accessibility to eldercare: Hansen method") +
  tm_layout(frame = TRUE) +
  tm_compass(type="8star", size = 2) +
  tm_scalebar() +
  tm_grid(alpha = 0.2)
```

```{r}

# Create maps for all power values for comparison
for (power_val in power_values) {
  col_name <- paste0("accHansen_power", power_val)
  temp_hexagon <- bind_cols(hexagons, hansen_results[[as.character(power_val)]])
  
  print(
    tm_shape(temp_hexagon,
             bbox = mapex) + 
      tm_polygons(fill = col_name, 
                  fill.scale = tm_scale_intervals(
                    style = "quantile",
                    n = 10,
                    values = "brewer.blues")) +
      tm_shape(eldercare) +
      tm_dots(size = 0.3,
              fill = "red",
              col = "black") +
      tm_title(paste("Hansen Accessibility - Power =", power_val)) +
      tm_layout(frame = TRUE) +
      tm_compass(type="8star", size = 2) +
      tm_scalebar() +
      tm_grid(alpha = 0.2)
  )
}
```

::: insightbox
Power = 0.5 Weak distance decay - accessibility spreads widely across Singapore, creating a smooth pattern with minimal variation between near and far locations.

Power = 1.0 Moderate decay - clearer concentration around facilities emerges, but surrounding areas still maintain reasonable accessibility with visible gradients.

Power = 1.5 Strong decay - steep accessibility drop-off creates distinct service zones around each facility, highlighting spatial inequality between central and peripheral areas.

Power = 2.0 Very strong decay - accessibility highly localized immediately around facilities, with large regions showing near-zero access and sharp peaks at facility locations.

Power = 2.5 Extreme decay - accessibility limited to immediate facility vicinity, vast majority of Singapore shows minimal access, emphasizing critical importance of proximity.

As the power parameter increases, accessibility becomes increasingly concentrated around eldercare facilities due to stronger distance decay. Lower powers (0.5-1.0) create more equitable but potentially overestimated accessibility, while higher powers (1.5-2.5) produce highly localized patterns that better reflect distance friction but may underestimate access for intermediate distances. The choice of power value represents a trade-off between modeling realistic distance barriers and capturing broader spatial accessibility patterns.
:::
