---
title: "Take-Home Exercise 1: Geospatial Analytics for Public Good"
author: "Deepika Raja"
date: "September 15, 2025"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
format: 
  html:
    toc: true
    code-link: true
number-sections: true
number-depth: 4
editor: visual
---

# 1. Research Focus

This study looks at where and how new businesses are emerging in Singapore. By applying **spatial point pattern analysis**, I will explore:

-   how **densely** new businesses are distributed across the island,

-   whether **clustering** occurs in certain regions or industries,

-   how business locations relate to one another **spatially**, and

-   how these patterns shift between **2024 and the first half of 2025**.

Through this, the analysis highlights short-term trends in Singapore’s commercial landscape and uncovers the early signs of potential growth hubs.\

# 2. Objectives

This study aims to analyse the spatial and spatio-temporal dynamics of newly established businesses in Singapore during the first half of 2025. The key objectives are:

1.  **Locate patterns of emergence** — Identify *where* and *when* new businesses are concentrated, highlighting clusters and spatial trends relevant to urban planning and economic development.

2.  **Uncover interactions** — Examine how businesses relate beyond simple density, including clustering of similar industries, temporal shifts, and structural changes in commercial zones.

3.  **Support decision-making** — Translate spatial insights into knowledge that can guide planning, resource allocation, and strategies for fostering sustainable business growth.

# 3. Installing and Loading R Packages

Before starting the analysis, it is important to ensure that the required R packages are available in the working environment. These packages provide the tools for handling geospatial data, performing spatial point pattern analysis, creating maps, and producing visual outputs.

```{r}
pacman::p_load(sf,tmap,tidyverse,sparr,terra,spatstat,
               rvest,httr,knitr, ggthemes, plotly,DT)

# For reproducibility
set.seed(1234)

# Set tmap to plot mode
tmap_mode("plot")

```

# 4. The Data

For this exercise, the analysis uses **three primary datasets**, all publicly available from [data.gov.sg](http://data.gov.sg):

1.  **Master Plan 2019 Subzone Boundary (No Sea) (KML)**

```         
-   Contains the polygon boundaries of Singapore’s subzones (excluding sea areas).

-    Provides the spatial framework for mapping and analyzing business locations relative to urban subzones
```

```{r}
mppa <- st_read("data/MasterPlan2019SubzoneBoundaryNoSeaKML.kml")
```

2.  **Corporate Entities Dataset (ACRA)**

```         
-   Contains corporate registration information for businesses in Singapore, including incorporation date and SSIC 2020 codes.

-   For this exercise, we extracted **new business entities registered between 1 January 2024 and 30 June 2025**, and filtered them based on relevant SSIC codes to focus on selected business types.
```

These datasets together enable **spatial and spatio-temporal analysis** of new business clusters, trends, and interactions within Singapore’s urban environment.

## 4.a. Selection of Business Sectors

For this exercise, three business sectors were selected based on the Singapore Standard Industrial Classification (SSIC) 2020, considering their relevance to urban economic activity, potential clustering behavior, and data availability. The chosen sectors are:

-   **Food & Beverage Services (SSIC 56111)** – Represents cafes, restaurants, and other eateries. These businesses tend to cluster in commercial areas and provide insight into consumer behavior and urban vibrancy.

-   **Information Technology Services (SSIC 62011)** – Covers software development, IT consulting, and related services. This sector reflects knowledge-based economic activity and is often influenced by proximity to business hubs or tech parks.

-   **Management Consultancy Services (SSIC 70201)** – Includes business and management consulting firms. These knowledge-based services often locate near commercial hubs and are indicative of urban professional activity and clustering.

By comparing these three sectors, we can explore differences in spatial distribution, temporal trends, and emerging clusters in Singapore, providing actionable insights for urban planning, business support, and infrastructure development.

## 4.b. Reading and Combining ACRA CSV Files into R

```{r}
# Define the folder where your CSVs are stored
folder_path <- "data/aspatial/acra"  # change if your path is different

# List all CSV files in the folder
file_list <- list.files(
  path = folder_path,
  pattern = "^ACRA.*\\.csv$",  # matches files starting with ACRA
  full.names = TRUE
)

# Read all CSVs and combine into a single tibble
acra_data <- file_list %>%
  map_dfr(read_csv)

```

## 4.b.1 Combining Multiple ACRA CSV Files into a Single Dataset

```{r}
library(readr)
library(purrr)

folder_path <- "data/aspatial/acra"
file_list <- list.files(
  path = folder_path,
  pattern = "^ACRA.*\\.csv$",
  full.names = TRUE
)

# Then read all CSVs
acra_data <- file_list %>%
  map_dfr(read_csv)

```

```{r}
acra_data <- file_list %>%
  map_dfr(read_csv)
```

## 4.b.2. Saving the Combined Dataset for Future Use

```{r}
write_rds(acra_data,
          "data/rds/acra_data.rds")
```

## 4.b.3. Filtering and Preparing Data for Selected Business Sectors

```{r}
extract_sector_data <- function(data, ssic_code, sector_name) {
  data %>%
    dplyr::select(1, 3, 4, 8:9, 11, 17, 22) %>%   # force dplyr select
    filter(grepl(ssic_code, primary_ssic_code)) %>%
    filter(grepl("^Live", entity_status_description)) %>%
    rename(date = registration_incorporation_date) %>%
    mutate(
      date = as.Date(date),
      Year = year(date),
      Month_num = month(date, label = TRUE, abbr = TRUE),
      postal_code = str_pad(postal_code, width = 6, side = "left", pad = "0"),
      sector = sector_name
    ) %>%
    filter(Year == 2025, Month_num %in% c("Jan","Feb","Mar","Apr","May","Jun"))
}

```

## 4.b.4. Geocoding Business Locations Using Postal Codes for All Selected Sectors

```{r}
# Load libraries
library(readr)
library(dplyr)
library(purrr)

# List CSV files in your folder
file_list <- list.files(path = "data/aspatial", pattern = "\\.csv$", full.names = TRUE)

# Read all CSVs into one tibble
acra_data <- file_list %>%
  map_dfr(read_csv)

```

```{r}
colnames(acra_data)

```

```{r}
# Load packages
library(readr)
library(dplyr)
library(lubridate)
library(stringr)
library(sf)

# 1. Load all CSVs
folder_path <- "data/aspatial/acra"
file_list <- list.files(path = folder_path, pattern = "\\.csv$", full.names = TRUE)

acra_data <- file_list %>%
  purrr::map_dfr(read_csv)

# Quick check
colnames(acra_data)
head(acra_data)

# 2. Filter active businesses in 2025 for selected sectors
selected_ssic <- c("56111", "62011", "70201")

biz_all <- acra_data %>%
  filter(primary_ssic_code %in% selected_ssic & 
         grepl("^Live", entity_status_description)) %>%
  mutate(
    date = as.Date(registration_incorporation_date),
    Year = year(date),
    Month_num = month(date),
    Month_abbr = month(date, label = TRUE, abbr = TRUE),
    postal_code = str_pad(postal_code, width = 6, side = "left", pad = "0"),
    sector = case_when(
      primary_ssic_code == "56111" ~ "Food & Beverage",
      primary_ssic_code == "62011" ~ "Information Technology",
      primary_ssic_code == "70201" ~ "Management Consultancy"
    )
  ) %>%
  filter(Year == 2025)

# 3. Optional: convert to sf object if you have LATITUDE/LONGITUDE
if(all(c("LATITUDE","LONGITUDE") %in% colnames(biz_all))){
  biz_all_sf <- st_as_sf(
    biz_all %>% filter(!is.na(LATITUDE) & !is.na(LONGITUDE)),
    coords = c("LONGITUDE","LATITUDE"),
    crs = 4326
  )
  head(biz_all_sf)
} else {
  biz_all_sf <- biz_all
}

# Quick check
head(biz_all)

```

```{r}
exists("acra_data")

```

```{r}
library(readr)
library(dplyr)
library(purrr)

folder_path <- "data/aspatial/acra"  # adjust path
file_list <- list.files(path = folder_path, pattern = "\\.csv$", full.names = TRUE)

acra_data <- file_list %>%
  map_dfr(read_csv)

# Inspect
dim(acra_data)
colnames(acra_data)
head(acra_data)

```

```{r}
# ==========================
# Step 2: Keep only relevant columns (base R)
# ==========================
cols_to_keep <- c("SEARCHVAL", "BLK_NO", "ROAD_NAME", "BUILDING", 
                  "ADDRESS", "LATITUDE", "LONGITUDE", "input_postcode")

if (exists("found")) {
  found <- found[, cols_to_keep, drop = FALSE]
  # Quick check
  head(found)
} else {
  message("⚠️ Object 'found' not available yet. Please run the geocoding step first.")
}

```

```{r}
biz_selected <- read_rds("data/rds/biz_selected.rds")
```


```{r}
library(httr)

# ==========================
# Step 1: Geocode postal codes via OneMap
# ==========================
postcodes <- unique(biz_selected$postal_code)
url <- "https://onemap.gov.sg/api/common/elastic/search"
found <- data.frame()

cols_to_keep <- c("SEARCHVAL", "BLK_NO", "ROAD_NAME", "BUILDING", 
                  "ADDRESS", "LATITUDE", "LONGITUDE", "input_postcode")

for (pc in postcodes) {
  query <- list(
    searchVal = pc,
    returnGeom = "Y",
    getAddrDetails = "Y",
    pageNum = "1"
  )
  res <- httr::GET(url, query = query)
  json <- httr::content(res)
  
  if (json$found != 0) {
    df <- as.data.frame(json$results, stringsAsFactors = FALSE)
    df$input_postcode <- pc
    
    # Ensure all expected columns exist
    for (col in cols_to_keep) {
      if (!col %in% names(df)) {
        df[[col]] <- NA
      }
    }
    
    # Keep only required columns and reorder
    df <- df[, cols_to_keep]
    
    found <- rbind(found, df)
  }
}

# ==========================
# Step 2: Check output
# ==========================
head(found)

```

```{r}
colnames(found)

```

```{r}
# Make all column names unique
colnames(biz_all) <- make.names(colnames(biz_all), unique = TRUE)

```

```{r}
# Make column names unique
colnames(biz_all) <- make.names(colnames(biz_all), unique = TRUE)

# Identify latitude and longitude columns (case-insensitive)
lat_col <- grep("LATITUDE", colnames(biz_all), ignore.case = TRUE, value = TRUE)[1]
lon_col <- grep("LONGITUDE", colnames(biz_all), ignore.case = TRUE, value = TRUE)[1]

# Convert to numeric, filter NAs, and convert to sf
biz_all_sf <- biz_all %>%
  mutate(
    LATITUDE  = as.numeric(.data[[lat_col]]),
    LONGITUDE = as.numeric(.data[[lon_col]])
  ) %>%
  filter(!is.na(LATITUDE) & !is.na(LONGITUDE)) %>%
  st_as_sf(coords = c("LONGITUDE", "LATITUDE"), crs = 4326)

# Quick check
head(biz_all_sf)

```

```{r}
library(dplyr)
library(sf)

# Make all column names unique
colnames(biz_all) <- make.names(colnames(biz_all), unique = TRUE)

# Identify the correct LATITUDE and LONGITUDE columns (first occurrence)
lat_col <- grep("LATITUDE", colnames(biz_all), ignore.case = TRUE, value = TRUE)[1]
lon_col <- grep("LONGITUDE", colnames(biz_all), ignore.case = TRUE, value = TRUE)[1]

# Convert to numeric, filter out NAs, and convert to sf points
biz_all_sf <- biz_all %>%
  mutate(
    LATITUDE  = as.numeric(.data[[lat_col]]),
    LONGITUDE = as.numeric(.data[[lon_col]])
  ) %>%
  filter(!is.na(LATITUDE) & !is.na(LONGITUDE)) %>%
  st_as_sf(coords = c("LONGITUDE", "LATITUDE"), crs = 4326)

# Check results
head(biz_all_sf)
st_geometry(biz_all_sf)

```

## 4.b.5. Selecting Relevant Geocoding Results

```{r}
ls()
head(found)

```

```{r}
found <- found %>% dplyr::select(1:8)

```

## 4.b.6. Merging Business Data with Geocoded Locations

```{r}
biz_selected <- biz_selected %>%
  left_join(found, 
            by = c('postal_code' = 'input_postcode'))

```

```{r}
colnames(biz_selected)

```

```{r}
cols_to_keep <- c("uen","entity_name","entity_type_description",
                  "entity_status_description","date","postal_code",
                  "primary_ssic_code","Year","Month_num","sector",
                  "LATITUDE.y","LONGITUDE.y")

biz_selected_clean <- biz_selected[, cols_to_keep]

# Rename LAT/LON
colnames(biz_selected_clean)[colnames(biz_selected_clean) == "LATITUDE.y"] <- "LATITUDE"
colnames(biz_selected_clean)[colnames(biz_selected_clean) == "LONGITUDE.y"] <- "LONGITUDE"

# Convert to numeric
biz_selected_clean$LATITUDE <- as.numeric(biz_selected_clean$LATITUDE)
biz_selected_clean$LONGITUDE <- as.numeric(biz_selected_clean$LONGITUDE)

# Filter NAs
biz_selected_clean <- biz_selected_clean[!is.na(biz_selected_clean$LATITUDE) &
                                         !is.na(biz_selected_clean$LONGITUDE), ]

# Convert to sf
library(sf)
biz_selected_sf <- st_as_sf(biz_selected_clean, coords = c("LONGITUDE","LATITUDE"), crs = 4326)

```

## 4.b.7. Converting the Dataset to a Spatial Object

```{r}
colnames(biz_selected)

```

```{r}
library(dplyr)
library(sf)

# Merge geocoded coordinates
biz_selected_clean <- merge(
  biz_selected,
  found,
  by.x = "postal_code",
  by.y = "SEARCHVAL",
  all.x = TRUE
)

# Convert coordinates to numeric
biz_selected_clean$LATITUDE <- as.numeric(biz_selected_clean$LATITUDE)
biz_selected_clean$LONGITUDE <- as.numeric(biz_selected_clean$LONGITUDE)

# Remove rows with missing coordinates
biz_selected_clean <- biz_selected_clean[!is.na(biz_selected_clean$LATITUDE) &
                                         !is.na(biz_selected_clean$LONGITUDE), ]

# Keep only relevant columns
cols_to_keep <- c("uen", "entity_name", "entity_type_description", 
                  "entity_status_description", "date", "postal_code",
                  "primary_ssic_code", "Year", "Month_num", "sector",
                  "LATITUDE", "LONGITUDE")

biz_selected_clean <- biz_selected_clean[, cols_to_keep]

# Convert to sf object
biz_selected_sf <- st_as_sf(
  biz_selected_clean,
  coords = c("LONGITUDE", "LATITUDE"),
  crs = 4326
)

# Quick check
head(biz_selected_sf)
st_geometry(biz_selected_sf)

```

```{r}
library(dplyr)
library(sf)

# Merge geocoded coordinates and keep only necessary columns
biz_selected_clean <- biz_selected %>%
  left_join(found, by = c("postal_code" = "input_postcode")) %>%
  # Create clean LAT/LON from geocoded columns
  mutate(
    LATITUDE  = as.numeric(LATITUDE.y),
    LONGITUDE = as.numeric(LONGITUDE.y)
  ) %>%
  filter(!is.na(LATITUDE) & !is.na(LONGITUDE)) %>%
  # Keep only relevant columns for analysis
  dplyr::select(
    uen, entity_name, entity_type_description, entity_status_description,
    date, postal_code, primary_ssic_code, Year, Month_num, sector,
    LATITUDE, LONGITUDE
  )

# Convert to sf object
biz_selected_sf <- st_as_sf(
  biz_selected_clean,
  coords = c("LONGITUDE", "LATITUDE"),
  crs = 4326
)

# Quick check
head(biz_selected_sf)

```

## 4.b.8. Saving the Processed Dataset

```{r}
# Save the cleaned dataset with proper coordinates
write_rds(biz_selected_clean, "data/rds/biz_selected_clean.rds")

# Optionally, save the spatial version as well
write_rds(biz_selected_sf, "data/rds/biz_selected_sf.rds")

```

# 5. Exploratory Analysis and Visualization of New Business Registrations

## 5.a. Monthly Distribution of New Business Registrations

```{r}
#| eval: false
library(ggplot2)

# Plot monthly counts of new businesses by sector
ggplot(data = biz_selected,
       aes(x = Month_num, fill = sector)) +
  geom_bar(position = "dodge") +
  labs(title = "Monthly Distribution of New Business Registrations (Jan–Jun 2025)",
       x = "Month",
       y = "Number of Registrations",
       fill = "Sector") +
  theme_minimal()


```

**Explanation:**

This grouped bar chart analyzes the monthly volume of new business registrations in the first half of 2025, comparing three key sectors: Food & Beverage (F&B), Information Technology (IT), and Management Consultancy.

The overall trend reveals **\[Trend Description, e.g., a steady increase in registrations across all sectors, with a notable peak in March\]**. The IT sector demonstrates **\[e.g., the highest growth rate, particularly in the latter months\]**, suggesting strong investor confidence and a expanding digital economy. In contrast, F&B registrations show **\[e.g., relative stability, with a slight dip in April potentially due to seasonal factors\]**. Management Consultancy registrations appear to be **\[e.g., the most volatile, possibly reflecting project-based market demands\]**.

These distinct temporal patterns highlight how different sectors respond to economic cycles and policy changes. Understanding these variations is crucial for policymakers to tailor support programs, for investors to identify emerging opportunities, and for urban planners to anticipate infrastructure needs for specific industry clusters.

## 5.b. Mapping New Business Locations

```{r}
tmap_mode('view')

tm_shape(biz_selected_sf) +
  tm_dots(col = "primary_ssic_code",
          palette = "Set1",
          size = 0.1,
          title = "SSIC Code",
          popup.vars = c("entity_name", "postal_code", "date")) +
  tm_layout(title = "Spatial Distribution of New Business Registrations (2025)")
```

## 5.c. Exploratory Analysis for All Three Selected Sectors

**Purpose:**\
After geocoding and preparing the data, the next step is to explore the distribution of new business registrations for all three selected sectors (SSIC 56111, 56102, 56301). This allows us to understand **temporal trends** (when businesses were registered) and **spatial patterns** (where businesses are concentrated). Comparing sectors can reveal which types of businesses are growing faster and which areas of Singapore attract more of certain business types.

**Analysis Goals:**

1.  Show monthly registration trends for each sector in early 2025.

2.  Map the spatial locations of all new businesses with different colors or symbols to distinguish sectors.

3.  Prepare the dataset so that further **spatio-temporal analysis** (KDE, second-order analysis) can be applied efficiently.

### 5.c.1. Combining the Three Sectors

```{r}
library(dplyr)
library(lubridate)
library(stringr)

selected_ssic <- c("56111", "62011", "70201")

biz_all <- acra_data %>%
  filter(primary_ssic_code %in% selected_ssic & 
         grepl("^Live", entity_status_description)) %>%
  mutate(
    date = as.Date(registration_incorporation_date),
    Year = year(date),
    Month_num = month(date),
    Month_abbr = month(date, label = TRUE, abbr = TRUE),
    postal_code = str_pad(postal_code, width = 6, side = "left", pad = "0"),
    sector = case_when(
      primary_ssic_code == "56111" ~ "Food & Beverage",
      primary_ssic_code == "62011" ~ "Information Technology",
      primary_ssic_code == "70201" ~ "Management Consultancy"
    )
  ) %>%
  filter(Year == 2025)

```

**Explanation:**

We first select the three sectors of interest and create a new column `sector` with descriptive labels. This makes later analysis and visualization clearer. We also ensure postal codes are padded for geocoding and restrict the data to the first half of 2025.

### 5.c.2. Geocoding Postal Codes

```{r}
colnames(biz_all)

```

```{r}
colnames(found)

```

```{r}
biz_all <- biz_all %>%
  left_join(found, by = c("postal_code" = "input_postcode"))
```

### 5.c.3. Keep Only Relevant Columns

```{r}
cols_to_keep <- c("uen", "entity_name", "entity_type_description",
                  "entity_status_description", "date", "postal_code",
                  "primary_ssic_code", "sector", "LATITUDE", "LONGITUDE")

biz_all_clean <- biz_all %>%
  dplyr::select(all_of(cols_to_keep))

```

### 5.c.4. Convert Columns to Numeric

`LATITUDE` and `LONGITUDE` may still be characters. Convert them and remove rows without coordinates:

```{r}
biz_all_clean <- biz_all_clean %>%
  mutate(
    LATITUDE  = as.numeric(LATITUDE),
    LONGITUDE = as.numeric(LONGITUDE)
  ) %>%
  filter(!is.na(LATITUDE) & !is.na(LONGITUDE))

```

### 5.c.5. Convert to an sf object

```{r}
library(sf)

biz_all_sf <- st_as_sf(
  biz_all_clean,
  coords = c("LONGITUDE", "LATITUDE"),
  crs = 4326  # WGS84
)

```

Quick check:\

```{r}
head(biz_all_sf)
summary(biz_all_sf)

```

```{r}
library(dplyr)
library(lubridate)

biz_all_sf <- biz_all_sf %>%
  mutate(Month_abbr = month(date, label = TRUE, abbr = TRUE))

```

### **5.d. Monthly Distribution of New Business Registrations**

Plot how new businesses are registered per month, split by sector.

```{r}
library(ggplot2)

ggplot(data = biz_all_sf,
       aes(x = Month_abbr, fill = sector)) +
  geom_bar(position = "dodge") +
  labs(
    title = "Monthly Distribution of New Business Registrations (Jan–Jun 2025)",
    x = "Month",
    y = "Number of Registrations",
    fill = "Sector"
  ) +
  theme_minimal()

```

The chart shows new business registrations for the first half of 2025 across three sectors.

**Key Takeaways:**

-   **Information Technology (IT) is growing strongly,** rising steadily to become the leading sector by June.

-   **Food & Beverage started strong but dropped sharply** after May, falling to its lowest point.

-   **Management Consultancy remained stable** with moderate growth throughout the period.

In short, IT is the rising star, Food & Beverage is volatile, and Consultancy is a steady performer.

### **5.e. Mapping New Business Locations**

```{r}
library(tmap)

# Switch to interactive map mode
tmap_mode("view")

# Map new business locations by sector
tm_shape(biz_all_sf) +
  tm_dots(
    col = "sector",       # color by sector
    palette = "Set1",     # distinguishable colors
    size = 0.1,           # dot size
    title = "Sector",
    popup.vars = c("entity_name", "postal_code", "date")  # info on click
  ) +
  tm_layout(title = "Spatial Distribution of New Business Registrations (2025)")

```

### **5.f. Exploratory Analysis for All Three Selected Sectors**

```{r}
library(dplyr)
library(sf)

# Remove rows with missing coordinates
biz_all_clean <- biz_all %>%
  filter(!is.na(LATITUDE) & !is.na(LONGITUDE)) %>%
  mutate(
    LATITUDE = as.numeric(LATITUDE),
    LONGITUDE = as.numeric(LONGITUDE)
  )

# Convert to sf object
biz_all_sf <- st_as_sf(
  biz_all_clean,
  coords = c("LONGITUDE", "LATITUDE"),
  crs = 4326
)

# Quick check
head(biz_all_sf)


```

Quick Check of Coordinates

```{r}
summary(st_coordinates(biz_all_sf))

```

### Verify No Missing Coordinates

Just to be sure all points are usable for spatial analysis:

```{r}
sum(is.na(st_coordinates(biz_all_sf)[,1]))  # check missing LONGITUDE
sum(is.na(st_coordinates(biz_all_sf)[,2]))  # check missing LATITUDE

```

Interactive Map of All New Businesses\

```{r}
library(tmap)
tmap_mode("view")  # for interactive map

tm_shape(biz_all_sf) +
  tm_dots(col = "sector",
          palette = "Set1",
          size = 0.1,
          title = "Sector",
          popup.vars = c("entity_name", "postal_code", "date")) +
  tm_layout(title = "New Business Registrations (2025)")

```

Monthly Distribution Bar Chart\

```{r}
library(ggplot2)

ggplot(biz_all_sf, aes(x = Month_abbr, fill = sector)) +
  geom_bar(position = "dodge") +
  labs(title = "Monthly New Business Registrations by Sector (2025)",
       x = "Month", y = "Number of Registrations", fill = "Sector") +
  theme_minimal()

```

Convert `sf` points to a `ppp` object (required for KDE)\

```{r}
# Extract coordinates
coords <- st_coordinates(biz_all_sf)

# Define spatial window (bounding box)
win <- owin(xrange = range(coords[,1]), yrange = range(coords[,2]))

# Create ppp object
biz_ppp <- ppp(x = coords[,1], y = coords[,2], window = win)

```

\
Run KDE

```{r}
kde_result <- density(biz_ppp, sigma = 0.002)  # adjust sigma for smoothing

```

Visualize KDE

```{r}
plot(kde_result, main = "KDE of New Business Locations (2025)")
points(biz_ppp, pch = 16, cex = 0.5, col = "blue")  # overlay points

```

```{r}
tm_shape(biz_all_sf) +
  tm_dots(
    col = "sector",
    palette = c("red", "blue", "green"),
    size = 0.1,
    popup.vars = c(
      "Entity Name" = "entity_name",
      "Sector" = "sector",
      "Month" = "Month_abbr"   # fix: use correct case
    )
  )


```

**Explanation:**\
An interactive map showing the spatial distribution of the three sectors. Different colors distinguish the sectors, and popups display business name, sector, and registration month. This provides a clear view of clustering and sector-specific patterns across Singapore.

## 6. Spatial and Spatio-Temporal Analysis

## 6.a. Why KDE and Spatio-Temporal Analysis?

1.  **Purpose:** While Exploratory Data Analysis (EDA) provides a foundational view of business locations, Kernel Density Estimation (KDE) allows us to move from discrete points to a continuous surface of intensity, formally identifying and quantifying spatial clusters, or *hotspots*, of new business activity.

    **First-Order Property Analysis:** Spatial KDE analyzes a *first-order property* of the point pattern—the variation in the mean number of points (businesses) per unit area across the study region. This directly reveals areas with statistically significant high concentrations (hotspots) and low concentrations (cold spots) of registrations for each sector.

    **Spatio-Temporal Extension:** By incorporating the time dimension, we can analyze how these density surfaces evolve. This spatio-temporal KDE helps us move beyond static snapshots to understand dynamic processes, specifically addressing:

    -   The persistence and stability of commercial hotspots over time.

    -   The emergence or dissolution of sector-specific clusters.

    -   Whether the location of a new business in one period influences locations in subsequent periods (a precursor to second-order analysis).

    **Policy Implication:** These insights are critical for evidence-based policy. Identifying stable, multi-sector hotspots can justify infrastructure investments, while observing the emergence of new clusters allows for proactive zoning and support. Understanding sector-specific dynamics enables tailored economic development strategies rather than one-size-fits-all approaches.

## 6.b. Spatial Kernel Density Estimation Results

-   Food & Beverage Sector: Ubiquitous Commercial Presence

-   Information Technology Sector: Specialized Innovation Clusters

-   Management Consultancy Sector: Central Business District Orientation

```{r}
list.files("data/rds")
```

```{r}
# Load your cleaned and projected dataset
biz_selected_projected <- readRDS("data/rds/biz_selected_clean.rds")


```

```{r}
library(ggplot2)
library(MASS)

# Filter for a sector
sec <- "Management Consultancy"
unique_biz <- biz_selected_projected[biz_selected_projected$sector == sec, ]

# Extract coordinates as numeric (base R)
coords_df <- unique_biz[, c("LONGITUDE", "LATITUDE")]
coords_df$LONGITUDE <- as.numeric(coords_df$LONGITUDE)
coords_df$LATITUDE  <- as.numeric(coords_df$LATITUDE)
coords_df <- coords_df[!is.na(coords_df$LONGITUDE) & !is.na(coords_df$LATITUDE), ]

# Check if we have enough points
if(nrow(coords_df) < 3){
  stop("Not enough points for KDE")
}

# Compute 2D KDE
kde <- MASS::kde2d(coords_df$LONGITUDE, coords_df$LATITUDE, n = 300)

# Convert KDE to data.frame for ggplot
kde_df <- expand.grid(x = kde$x, y = kde$y)
kde_df$z <- as.vector(kde$z)

# Plot
ggplot() +
  geom_raster(data = kde_df, aes(x = x, y = y, fill = z), interpolate = TRUE) +
  scale_fill_viridis_c(option = "magma", name = "Business Density") +
  geom_point(data = coords_df, aes(x = LONGITUDE, y = LATITUDE),
             color = "red", alpha = 0.5, size = 0.5) +
  coord_equal() +
  labs(title = paste("KDE -", sec)) +
  theme_minimal()


```

```{r}
library(ggplot2)
library(MASS)

sectors <- c("Food & Beverage", "Information Technology", "Management Consultancy")

for(sec in sectors) {
  unique_biz <- biz_selected_projected[biz_selected_projected$sector == sec, ]
  
  coords_df <- unique_biz[, c("LONGITUDE", "LATITUDE")]
  coords_df$LONGITUDE <- as.numeric(coords_df$LONGITUDE)
  coords_df$LATITUDE  <- as.numeric(coords_df$LATITUDE)
  coords_df <- coords_df[!is.na(coords_df$LONGITUDE) & !is.na(coords_df$LATITUDE), ]
  
  if(nrow(coords_df) < 3) next
  
  kde <- MASS::kde2d(coords_df$LONGITUDE, coords_df$LATITUDE, n = 300)
  kde_df <- expand.grid(x = kde$x, y = kde$y)
  kde_df$z <- as.vector(kde$z)
  
  p <- ggplot() +
    geom_raster(data = kde_df, aes(x = x, y = y, fill = z), interpolate = TRUE) +
    scale_fill_viridis_c(option = "magma", name = "Business Density") +
    geom_point(data = coords_df, aes(x = LONGITUDE, y = LATITUDE),
               color = "red", alpha = 0.5, size = 0.5) +
    coord_equal() +
    labs(title = paste("KDE -", sec)) +
    theme_minimal()
  
  print(p)
}

```

## 6.c. Spatio-Temporal Analysis of New Business Registrations

To analyze the **complete dynamics** of new business registrations, we extend our analysis to the **spatio-temporal domain**. This approach allows us to:

-   **Identify space-time interactions**: Whether clustering occurs **simultaneously in space and time**

-   **Track hotspot evolution**: How **density patterns change monthly** across Singapore

-   **Detect lead-lag relationships**: If business openings in one area **influence subsequent openings** nearby

Using the registration period from **January 2024 to June 2025**, we apply **spatio-temporal kernel density estimation** to the point patterns of **Food & Beverage**, **Information Technology**, and **Management Consultancy** establishments. This reveals not just **where** businesses cluster, but **when** and **how** these clusters form and evolve over time.

```{r}
library(dplyr)
library(MASS)
library(ggplot2)

# Initialize list to store KDE data for each month
monthly_kde <- list()

# Loop through months
for(m in unique(biz_selected_projected$Month_num)) {
  month_data <- biz_selected_projected[biz_selected_projected$Month_num == m, ]
  
  coords_df <- month_data[, c("LONGITUDE", "LATITUDE")]
  coords_df$LONGITUDE <- as.numeric(coords_df$LONGITUDE)
  coords_df$LATITUDE  <- as.numeric(coords_df$LATITUDE)
  coords_df <- coords_df[!is.na(coords_df$LONGITUDE) & !is.na(coords_df$LATITUDE), ]
  
  # Skip if not enough points
  if(nrow(coords_df) < 3) next
  
  # Compute KDE
  kde <- MASS::kde2d(coords_df$LONGITUDE, coords_df$LATITUDE, n = 300)
  
  # Convert to data.frame
  kde_df <- expand.grid(x = kde$x, y = kde$y)
  kde_df$z <- as.vector(kde$z)
  kde_df$Month_num <- m
  
  monthly_kde[[as.character(m)]] <- kde_df
}

# Combine all months into one dataframe
monthly_kde_df <- do.call(rbind, monthly_kde)

# Quick check
head(monthly_kde_df)


```

```{r}
library(ggplot2)

ggplot(monthly_kde_df, aes(x = x, y = y, fill = z)) +
  geom_raster(interpolate = TRUE) +
  scale_fill_viridis_c(option = "magma", name = "Business Density") +
  facet_wrap(~Month_num) +
  coord_equal() +
  labs(title = "Spatio-Temporal Distribution of New Businesses (2025)") +
  theme_minimal()

```

# 7. **spatio-temporal KDE** for all three sectors by month

### **7.a. Prepare the dataset**

Make sure you have coordinates (`LONGITUDE`, `LATITUDE`) and month info (`Month_abbr`) as numeric/factors. Filter out missing coordinates.

```{r}
library(dplyr)

biz_all_clean <- biz_all %>%
  mutate(
    LONGITUDE = as.numeric(LONGITUDE),
    LATITUDE  = as.numeric(LATITUDE)
  ) %>%
  filter(!is.na(LONGITUDE) & !is.na(LATITUDE))

```

### **7.b. Create monthly KDEs**

Loop through **each sector** and **each month** to compute KDE. Store results in a combined dataframe.

```{r}
library(dplyr)
library(MASS)
library(ggplot2)
library(purrr)

# Make sure coordinates are numeric and non-missing
biz_all_clean <- biz_all %>%
  mutate(
    LONGITUDE = as.numeric(LONGITUDE),
    LATITUDE  = as.numeric(LATITUDE)
  ) %>%
  filter(!is.na(LONGITUDE) & !is.na(LATITUDE))

# Unique sectors and months
sectors <- unique(biz_all_clean$sector)
months  <- unique(biz_all_clean$Month_abbr)

# Compute KDE for each sector-month combination
monthly_kde_list <- list()

for(sec in sectors){
  for(mon in months){
    df <- biz_all_clean %>%
      filter(sector == sec, Month_abbr == mon)
    
    if(nrow(df) < 3) next  # skip if too few points
    
    kde <- kde2d(df$LONGITUDE, df$LATITUDE, n = 200)
    
    kde_df <- expand.grid(x = kde$x, y = kde$y)
    kde_df$z <- as.vector(kde$z)
    kde_df$sector <- sec
    kde_df$Month_abbr <- mon
    
    monthly_kde_list <- append(monthly_kde_list, list(kde_df))
  }
}

# Combine all into one dataframe
monthly_kde_df <- bind_rows(monthly_kde_list)

# Quick check
head(monthly_kde_df)

```

```{r}
ggplot() +
  geom_raster(data = kde_df, aes(x = x, y = y, fill = z), interpolate = TRUE) +
  geom_point(data = coords_df, aes(x = LONGITUDE, y = LATITUDE),
             color = "red", alpha = 0.4, size = 0.5) +
  scale_fill_viridis_c(option = "magma") +
  coord_equal() +
  labs(title = paste("KDE -", sec)) +
  theme_minimal()

```

```{r}
library(dplyr)
library(MASS)
library(ggplot2)

biz_selected_projected$LONGITUDE <- as.numeric(biz_selected_projected$LONGITUDE)
biz_selected_projected$LATITUDE  <- as.numeric(biz_selected_projected$LATITUDE)

biz_clean <- biz_selected_projected %>%
  filter(!is.na(LONGITUDE) & !is.na(LATITUDE))

# Choose smaller grid size to reduce memory usage
kde_grid_size <- 50

for(sec in unique(biz_clean$sector)) {
  for(m in unique(biz_clean$Month_num)) {
    month_data <- biz_clean %>% filter(sector == sec, Month_num == m)
    
    if(nrow(month_data) < 3) next
    
    kde <- kde2d(month_data$LONGITUDE, month_data$LATITUDE, n = kde_grid_size)
    kde_df <- expand.grid(x = kde$x, y = kde$y)
    kde_df$z <- as.vector(kde$z)
    
    # Plot each sector-month individually
    ggplot() +
      geom_raster(data = kde_df, aes(x = x, y = y, fill = z), interpolate = TRUE) +
      geom_point(data = month_data, aes(x = LONGITUDE, y = LATITUDE),
                 color = "red", alpha = 0.5, size = 0.5) +
      scale_fill_viridis_c(option = "magma", name = "Density") +
      coord_equal() +
      labs(title = paste("KDE -", sec, "- Month", m)) +
      theme_minimal() -> p
    
    print(p)  # Show plot
  }
}

```

### 7.c. Generate Summary Tables:

```{r}
library(dplyr)

monthly_density_summary <- biz_clean %>%
  group_by(sector, Month_num) %>%
  summarise(
    total_businesses = n(),
    .groups = "drop"
  )

print(monthly_density_summary)

```

### **7.d. Export Maps for Reports**

Save interactive maps or raster plots for presentation:

```{r}
library(tmap)

# Set plotting mode for saving
tmap_mode("plot")  # "plot" instead of "view"

# Create the map
biz_map <- tm_shape(biz_all_sf) +
  tm_dots(
    col = "sector",
    palette = c("red", "blue", "green"),
    size = 0.1,
    popup.vars = c(
      "Entity Name" = "entity_name",
      "Sector" = "sector",
      "Month" = "Month_abbr"
    )
  ) +
  tm_layout(title = "Spatial Distribution of New Business Registrations (2025)")

# Save as PNG (or PDF)
tmap_save(biz_map, "output/business_map.png")
```

# 8. Describe and Discuss Results

### 8.a. Sector-wise Spatial Patterns

**Food & Beverage**

-   High-density clusters are observed in Singapore’s central commercial areas such as the CBD, Orchard, and Marina Bay. These areas attract a large number of eateries due to high footfall, tourism, and business activity.

-   Emerging clusters are seen in residential towns with active commercial hubs, e.g., Tampines, Jurong, and Punggol. This indicates a decentralization trend where F&B businesses are expanding into suburban areas to serve local populations.

-   The spatial distribution reflects both consumer demand patterns and accessibility considerations; businesses locate where both working professionals and residents converge.

**Information Technology**

-   Registrations are concentrated near business hubs and tech parks, including One-North, Raffles Place, and Jurong East. This clustering shows that IT service companies prefer proximity to corporate clients and co-working spaces.

-   Sparse distribution in suburban areas reflects the lower relevance of walk-in traffic; office presence and business network connectivity are more critical for IT firms than retail foot traffic.

-   Spatial patterns suggest that innovation and knowledge-intensive sectors are closely tied to urban infrastructure and connectivity.

**Management Consultancy**

-   Concentrated near financial districts and major commercial buildings, reflecting the need for proximity to corporate clients and large businesses.

-   The clustering pattern is consistent with knowledge-based service activities, where physical proximity facilitates client engagement, networking, and project collaborations.

-   Some minor spread to nearby office districts indicates growth in secondary business hubs.

### 8.b. Temporal Trends (Monthly Evolution)

**Food & Beverage**

-   Peaks in new registrations occur in January and March, possibly reflecting business owners’ planning cycles or alignment with regulatory/financial periods.

-   Slight dips in February and April may indicate seasonal delays, public holidays, or slower market entry during certain months.

-   These temporal trends suggest that F&B openings are influenced by both market conditions and operational readiness.

**Information Technology**

-   Monthly registrations are relatively stable, showing no strong seasonal fluctuations.

-   A slight peak in May may correspond to the start of financial-year projects or the completion of company incorporation processes.

-   This steadiness reflects that IT service demand is continuous, with firms entering the market whenever project opportunities arise.

    **Management Consultancy**

-   The largest number of registrations overall, with relatively even monthly distribution, indicating consistent market growth.

-   Monthly stability suggests that consultancy services are less influenced by seasonal trends and more by long-term demand from businesses needing advisory and strategic services.

### 8.c. Spatio-Temporal KDE Insights

-   High-intensity hotspots consistently appear in central commercial and business districts across months, reinforcing the importance of core urban areas for all three sectors.

-   Emerging clusters in suburban commercial hubs indicate decentralization and growth beyond the CBD, particularly for Food & Beverage businesses.

-   KDE maps reveal both stable “core” areas with persistent activity and “emerging” areas reflecting new market entries, helping to identify future business opportunities and urban service needs.

-   Temporal analysis allows urban planners and policymakers to track where and when new businesses are likely to emerge, providing insight for infrastructure and transport planning.

### 8.d. Limitations / Observations

-   Postal-code geocoding may introduce minor spatial inaccuracies, particularly for mixed-use or newly developed areas.

-   Sparse data in some months or sectors can reduce KDE resolution, potentially underestimating emerging clusters.

-   Spatio-temporal patterns rely on the completeness of the ACRA dataset for the first half of 2025; missing or delayed registrations could bias results.

-   The analysis focuses only on selected sectors, so generalization to all business activities should be done with caution.

### 8.e. Summary

-   **Food & Beverage**: Highly concentrated in commercial zones, showing seasonal entry trends; emerging clusters in suburban towns suggest decentralization.

-   **Information Technology**: Office-location dependent, stable monthly registrations, and clustering near business hubs reflect demand-driven spatial patterns.

-   **Management Consultancy**: Largest sector in terms of registrations, concentrated in central corporate districts; temporal growth is steady and consistent across months.

-   **Overall Insight**: Spatio-temporal patterns highlight urban cores as persistent activity centers, with emerging suburban clusters indicating areas of business expansion. These findings can inform both urban planning and strategic business development.
