---
title: "Take-Home Exercise 1: Geospatial Analytics for Public Good"
author: "Deepika Raja"
date: "September 15, 2025"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
format: 
  html:
    toc: true
    code-link: true
number-sections: true
number-depth: 4
editor: visual
---

# 1. Research Focus

This study looks at where and how new businesses are emerging in Singapore. By applying **spatial point pattern analysis**, I will explore:

-   how **densely** new businesses are distributed across the island,

-   whether **clustering** occurs in certain regions or industries,

-   how business locations relate to one another **spatially**, and

-   how these patterns shift between **2024 and the first half of 2025**.

Through this, the analysis highlights short-term trends in Singapore’s commercial landscape and uncovers the early signs of potential growth hubs.

# 2. Objectives

This study aims to analyse the spatial and spatio-temporal dynamics of newly established businesses in Singapore during the first half of 2025. The key objectives are:

1.  **Locate patterns of emergence** — Identify *where* and *when* new businesses are concentrated, highlighting clusters and spatial trends relevant to urban planning and economic development.

2.  **Uncover interactions** — Examine how businesses relate beyond simple density, including clustering of similar industries, temporal shifts, and structural changes in commercial zones.

3.  **Support decision-making** — Translate spatial insights into knowledge that can guide planning, resource allocation, and strategies for fostering sustainable business growth.

# 3. Installing and Loading R Packages

Before starting the analysis, it is important to ensure that the required R packages are available in the working environment. These packages provide the tools for handling geospatial data, performing spatial point pattern analysis, creating maps, and producing visual outputs.

```{r}
pacman::p_load(sf,tmap,tidyverse,sparr,terra,spatstat,
               rvest,httr,knitr, ggthemes, plotly,DT)

# For reproducibility
set.seed(1234)

# Set tmap to plot mode
tmap_mode("plot")
```

# 4. The Data

For this exercise, the analysis uses **three primary datasets**, all publicly available from [data.gov.sg](http://data.gov.sg):

## 4.a. Master Plan 2019 Subzone Boundary (No Sea) (KML)

-   Contains the polygon boundaries of Singapore’s subzones (excluding sea areas).
-   Provides the spatial framework for mapping and analyzing business locations relative to urban subzones.

```{r}
mppa <- st_read("data/MasterPlan2019SubzoneBoundaryNoSeaKML.kml")
```

The KML file `MasterPlan2019SubzoneBoundaryNoSeaKML.kml` was successfully loaded using the KML driver. It contains a simple feature collection of 332 multipolygon features with two attribute fields. Each feature corresponds to a URA Master Plan 2019 subzone (excluding sea areas) and is defined using two-dimensional geographic coordinates (longitude and latitude) under the WGS 84 coordinate system. The dataset spans the entirety of Singapore, with a bounding box approximately from 103.61°E to 104.09°E longitude and 1.16°N to 1.47°N latitude.

Next, we will extract the relevant information from the HTML source.

```{r}
extract_kml_field <- function(html_text, field_name) {
  if (is.na(html_text) || html_text == "") return(NA_character_) # check if text is na
  
  page <- read_html(html_text)
  rows <- page %>% html_elements("tr") # find table row element 
  
  value <- rows %>%
    # looks through each row to find table header == field_name and
    keep(~ html_text2(html_element(.x, "th")) == field_name) %>% 
    html_element("td") %>% # extract table data
    html_text2() # and then extract clean text context
  
  if (length(value) == 0) NA_character_ else value # return value if is not NA
}


```

```{r}
mppa <- mppa %>%
  mutate(REGION_N = map_chr(Description,extract_kml_field,"REGION_N"),
         PLN_AREA_N = map_chr(Description,extract_kml_field,"PLN_AREA_N"))%>%
  select(-Name, -Description) %>% # unselect columns "Name" and "Description"
  relocate(geometry, .after = last_col()) # move the "geometry" to the end

```

We transform the dataset’s coordinate reference system (CRS) from WGS84 (EPSG:4326) to SVY21 (EPSG:3414) to enhance accuracy and support further spatial analyses. This conversion changes all feature coordinates from (latitude, longitude) in degrees to (x, y) in meters within Singapore’s SVY21 system.

```{r}
mppa <- mppa %>%
    #Removes Z (elevation) and M (measure) dimensions from the geometry if they exist
  st_zm(drop = TRUE, what = "ZM") %>% 
  st_transform(crs = 3414)
st_geometry(mppa)

```

```{r}
write_rds(mppa,"data/rds/mppa.rds")

```

```{r}
mppa <- read_rds("data/rds/mppa.rds")

```

### 4.a.1. Let's plot planning area in SG:

```{r}
#| fig.width: 12
#| fig.height: 10

tm_shape(mppa) +
  tm_polygons(
      col = "REGION_N",
      palette = "Set3",  # updated color palette
      lwd = 0.5
  ) +
  tm_text(
      "PLN_AREA_N", size = 0.4,
      auto.placement = TRUE,
      remove.overlap = TRUE
  ) +
  tm_title(
      "SG Planning Area",
      position = tm_pos_out("center", "top", "center")
  ) +
  tm_legend(
      frame = FALSE,
      position = c("right", "bottom")
  ) +
  tm_layout(
      frame = FALSE,
      bg.color = "#f6f6f6"
  )

```

```{r}
mppa_sf <- mppa %>%
  filter (PLN_AREA_N != "CENTRAL WATER CATCHMENT",
          PLN_AREA_N != "WESTERN WATER CATCHMENT",
         PLN_AREA_N != "WESTERN ISLANDS",
         PLN_AREA_N != "SOUTHERN ISLANDS",
         PLN_AREA_N != "NORTH-EASTERN ISLANDS",
         )

```

We focus on Singapore’s main island, excluding non-commercial areas, to accurately show business-related spatial patterns. Now, we plot the area again :\

```{r}
#| fig.width: 12
#| fig.height: 10

tm_shape(mppa_sf) +
  tm_polygons(col ="REGION_N",
              palette = "Pastel2",
              lwd = 0.5) +
  tm_text("PLN_AREA_N", size=0.4,
          auto.placement = TRUE,
          remove.overlap = TRUE) +
  tm_title("SG Planning Area - Business Activities",
           position = tm_pos_out("center", "top", "center")) +
  tm_legend(frame = FALSE,
            position = c("right", "bottom")) +
  tm_layout(frame = FALSE,
            bg.color = "#e9f4f8")

```

## 4.b. Corporate Entities Dataset (ACRA)

-   Contains corporate registration information for businesses in Singapore, including incorporation date and SSIC 2020 codes.

-   For this exercise, we extracted **new business entities registered between 1 January 2024 and 30 June 2025**, and filtered them based on relevant SSIC codes to focus on selected business types.

These datasets together enable **spatial and spatio-temporal analysis** of new business clusters, trends, and interactions within Singapore’s urban environment.

```{r}

folder_path <- "data/ACRA"
file_list <- list.files(path = folder_path,
                        pattern = "^ACRA*.*\\.csv$",
                        full.names = TRUE)
acra <- file_list %>%
  map_dfr(read_csv)
```

### 4.b.1 Filtering SSIC Data:

Extract selected data for three industries: Food and Beverages (SSIC 56111), Management Consultancy (SSIC 70201), and Information Technology (SSIC 62011) from January to June 2025:

```{}
```

## 4.c. Geocoding Business Locations

The ACRA dataset does not include latitude and longitude; it only provides postal codes and addresses. To map these locations, we need **geocoding**—converting textual location information into precise geographic coordinates.

For this project, we use **OneMap**, Singapore’s official geospatial platform, to geocode each postal code. The API returns accurate coordinates, enabling us to plot business locations on a map and conduct spatial analyses such as clustering, density estimation, and proximity calculations. This step is crucial because spatial operations require numerical coordinates rather than just addresses.

### 4.c.1. Combining the data by row:

```{}
```

### 4.c.2. Saving Combined Dataset

### 4.c.3. Extracting Unique postal codes from extracted data:

```{}
```

There are 1925 unique observations.

Next, we utilize the **httr** package to make a GET request to the OneMap API in order to obtain geographic coordinates:
