---
title: "Hands on Exercise 4- Spatial Weights and Applications"
author: "Deepika Raja"
date: "September 19, 2025"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# **1 Overview**

In this hands-on exercise, we will learn how to compute spatial weights using R. We will learn to compute spatial weights, visualize spatial distributions, and create spatially lagged variables using various functions from R packages such as sf,spdep, and tmap

# **2 The Data**

We will use following geospatial datasets in this exercise:

| Dataset | Description | Format |
|-------------------|----------------------------------|-------------------|
| *Hunan* | Hunan county boundary layer geospatial data | ESRI shapefile |
| *Hunan_2012* | Contains selected Hunan’s local development indicators in 2012 | CSV file |

# **3 Installing and launching the R packages**

```{r}
pacman::p_load(sf, spdep, tmap, tidyverse, knitr)
```

# **4 Getting the Data Into R Environment**

### **4.a. Import shapefile into r environment**

The code chunk below uses [*st_read()*](https://r-spatial.github.io/sf/reference/st_read.html) of **sf** package to import Hunan shapefile into R. The imported shapefile will be **simple features** Object of **sf**.

```{r}
hunan <- st_read(dsn = "data/geospatial", 
                 layer = "Hunan")
```

### **4.b. Import csv file into r environment**

Next, we will import *Hunan_2012.csv* into R by using *read_csv()* of **readr** package. The output is R dataframe class.

```{r}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")
```

### **4.c. Performing relational join**

The code chunk below will be used to update the attribute table of *hunan*’s SpatialPolygonsDataFrame with the attribute fields of *hunan2012* dataframe. This is performed by using *left_join()* of **dplyr** package.

```{r}
hunan <- left_join(hunan,hunan2012) %>%
  select(1:4, 7, 15)
```

# **5. Visualising Regional Development Indicator**

Now, we are going to prepare a basemap and a choropleth map showing the distribution of GDPPC 2012 by using *qtm()* of **tmap** package.

```{r}
basemap <- tm_shape(hunan) +
  tm_polygons() +
  tm_text("NAME_3", size=0.5)

gdppc <- qtm(hunan, "GDPPC") +
  tm_layout(
    legend.position = c("left", "bottom"))
tmap_arrange(basemap, gdppc, asp=1, ncol=2)
```

# **6. Computing Contiguity Spatial Weights**

### **6.a. Computing (QUEEN) contiguity based neighbours**

The code chunk below is used to compute Queen contiguity weight matrix:

```{r}
wm_q <- poly2nb(hunan, queen=TRUE)
summary(wm_q)
```

The summary report above shows that there are 88 area units in Hunan. The most connected area unit has 11 neighbours. There are two area units with only one heighbours.

For each polygon in our polygon object, *wm_q* lists all neighboring polygons. For example, to see the neighbors for the first polygon in the object, type:

```{r}
wm_q[[1]]
```

Polygon 1 has 5 neighbors. The numbers represent the polygon IDs as stored in hunan SpatialPolygonsDataFrame class.

We can retrive the county name of Polygon ID=1 by using the code chunk below:

```{r}
hunan$County[1]
```

The output reveals that Polygon ID=1 is Anxiang county.

To reveal the county names of the five neighboring polygons, the code chunk will be used:

```{r}
hunan$NAME_3[c(2,3,4,57,85)]
```

We can retrieve the GDPPC of these five countries by using the code chunk below.

```{r}
nb1 <- wm_q[[1]]
nb1 <- hunan$GDPPC[nb1]
nb1
```

The printed output above shows that the GDPPC of the five nearest neighbours based on Queen’s method are 20981, 34592, 24473, 21311 and 22879 respectively.

You can display the complete weight matrix by using *str()*:

```{r}
str(wm_q)
```

Warning: The output might cut across several pages. Save the trees if you are going to print out the report.

### **6.b. Creating (ROOK) contiguity based neighbors**

The code chunk below is used to compute Rook contiguity weight matrix:

```{r}
wm_r <- poly2nb(hunan, queen=FALSE)
summary(wm_r)
```

The summary report above shows that there are 88 area units in Hunan. The most connect area unit has 10 neighbours. There are two area units with only one heighbours.

### **6.c. Visualising Contiguity Weights**

A connectivity graph links each point to its neighbors. Since we are working with **polygons**, we first need to extract **centroids** (points). This is done using the **sf** package.

To prepare for the graph:

-   Compute **polygon centroids** with `st_centroid`.

-   Extract **longitude** and **latitude** into a separate data frame.

    Use **`map_dbl`** from the *purrr* package to apply `st_centroid` across geometries and extract coordinates.

    -   `[[1]]` gives **longitude** (first coordinate).

    -   `[[2]]` gives **latitude** (second coordinate).

```{r}
longitude <- map_dbl(hunan$geometry, ~st_centroid(.x)[[1]])
```

We do the same for latitude with one key difference. We access the second value per each centroid with \[\[2\]\].

```{r}
latitude <- map_dbl(hunan$geometry, ~st_centroid(.x)[[2]])
```

Now that we have latitude and longitude, we use cbind to put longitude and latitude into the same object.

```{r}
coords <- cbind(longitude, latitude)
```

We check the first few observations to see if things are formatted correctly.

```{r}
head(coords)
```

#### 6.c.1. Plotting Queen contiguity based neighbours map

```{r}
plot(hunan$geometry, border="lightgrey")
plot(wm_q, coords, pch = 19, cex = 0.6, add = TRUE, col= "red")
```

#### 6.c.2. Plotting Rook contiguity based neighbours map

```{r}
plot(hunan$geometry, border="lightgrey")
plot(wm_r, coords, pch = 19, cex = 0.6, add = TRUE, col = "red")
```

#### 6.c.3. Plotting both Queen and Rook contiguity based neighbours maps

```{r}
par(mfrow=c(1,2))
plot(hunan$geometry, border="lightgrey", main="Queen Contiguity")
plot(wm_q, coords, pch = 19, cex = 0.6, add = TRUE, col= "red")
plot(hunan$geometry, border="lightgrey", main="Rook Contiguity")
plot(wm_r, coords, pch = 19, cex = 0.6, add = TRUE, col = "red")
```

## **7. Computing distance based neighbours**

In this section, you will learn how to create **distance-based weight matrices** using **`dnearneigh()`** from the *spdep* package.

-   The function identifies **neighbors** based on **Euclidean distance** within a specified range (`d1`, `d2`).

-   With **unprojected coordinates** (`longlat = TRUE`), it calculates **great-circle distances** (km) using the **WGS84 ellipsoid**.

### **7.a. Determine the Cut-off Distance**

To find the upper limit for the distance band:

1.  Use **`knearneigh()`** (spdep) to return the k-nearest neighbors of each point.

2.  Convert the result into a neighbors list with **`knn2nb()`**.

3.  Compute edge lengths with **`nbdists()`** (units depend on projection: coordinate units if projected, km if not).

4.  Flatten the list structure using **`unlist()`**.

```{r}
#coords <- coordinates(hunan)
k1 <- knn2nb(knearneigh(coords))
k1dists <- unlist(nbdists(k1, coords, longlat = TRUE))
summary(k1dists)
```

The summary report shows that the largest first nearest neighbour distance is 61.79 km, so using this as the upper threshold gives certainty that all units will have at least one neighbour.

### **7.b. Computing fixed distance weight matrix**

Now, we will compute the distance weight matrix by using *dnearneigh()* as shown in the code chunk below.

```{r}
wm_d62 <- dnearneigh(coords, 0, 62, longlat = TRUE)
wm_d62
```

Next, we will use *str()* to display the content of wm_d62 weight matrix.

```{r}
str(wm_d62)
```

Another way to display the structure of the weight matrix is to combine [*table()*](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/table) and [*card()*](https://r-spatial.github.io/spdep/reference/card.html) of spdep.\

```{r}
table(hunan$County, card(wm_d62))
```

```{r}
n_comp <- n.comp.nb(wm_d62)
n_comp$nc
```

```{r}
table(n_comp$comp.id)
```

#### 7.b.1. Plotting fixed distance weight matrix

Next, we will plot the distance weight matrix by using the code chunk below:

```{r}
plot(hunan$geometry, border="lightgrey")
plot(wm_d62, coords, add=TRUE)
plot(k1, coords, add=TRUE, col="red", length=0.08)
```

The red lines show the links of 1st nearest neighbours and the black lines show the links of neighbours within the cut-off distance of 62km.

Alternatively, we can plot both of them next to each other by using the code chunk below:

```{r}
par(mfrow=c(1,2))
plot(hunan$geometry, border="lightgrey", main="1st nearest neighbours")
plot(k1, coords, add=TRUE, col="red", length=0.08)
plot(hunan$geometry, border="lightgrey", main="Distance link")
plot(wm_d62, coords, add=TRUE, pch = 19, cex = 0.6)
```

### **7.c. Computing Adaptive Distance Weight Matrix**

In a **fixed distance weight matrix**, urban areas often have **more neighbors**, while rural areas have **fewer**. This can overly smooth neighbor relationships.

To address this, we can use **k-nearest neighbors (k-NN)** to directly control the number of neighbors, allowing either **asymmetric** or **symmetric** relationships.

```{r}
knn6 <- knn2nb(knearneigh(coords, k=6))
knn6
```

Similarly, we can display the content of the matrix by using *str()*.

```{r}
str(knn6)
```

Notice that each county has six neighbours, no less no more!

#### 7.c.1. Plotting distance based neighbours

We can plot the weight matrix using the code chunk below:

```{r}
plot(hunan$geometry, border="lightgrey")
plot(knn6, coords, pch = 19, cex = 0.6, add = TRUE, col = "red")
```

## **8. Weights based on IDW**

In this section, you will learn how to derive a spatial weight matrix based on Inversed Distance method.

First, we will compute the distances between areas by using [*nbdists()*](https://r-spatial.github.io/spdep/reference/nbdists.html) of **spdep**.

```{r}
dist <- nbdists(wm_q, coords, longlat = TRUE)
ids <- lapply(dist, function(x) 1/(x))
ids
```

# **9. Row-standardised Weights Matrix**

We assign equal weights to each neighboring polygon using style = "W". Each neighbor receives a weight of 1 / (# of neighbors), and the weighted values are summed.

-   Advantage: simple and intuitive.

-   Drawback: polygons at the edges have fewer neighbors, which may distort spatial autocorrelation.

Alternative: style = "B" (binary weights), among others.

```{r}
rswm_q <- nb2listw(wm_q, style="W", zero.policy = TRUE)
rswm_q
```

To see the weight of the first polygon’s eight neighbors type:

```{r}
rswm_q$weights[10]
```

Each neighbor is assigned a 0.125 of the total weight. This means that when R computes the average neighboring income values, each neighbor’s income will be multiplied by 0.125 before being tallied.

Using the same method, we can also derive a row standardised distance weight matrix by using the code chunk below:

```{r}
rswm_ids <- nb2listw(wm_q, glist=ids, style="B", zero.policy=TRUE)
rswm_ids
```

```{r}
rswm_ids$weights[1]
```

```{r}
summary(unlist(rswm_ids$weights))
```

## **10. Application of Spatial Weight Matrix**

In this section, you will learn how to create four different spatial lagged variables, they are:

-   spatial lag with row-standardized weights,

-   spatial lag as a sum of neighbouring values,

-   spatial window average, and

-   spatial window sum.

### **10.a. Spatial lag with row-standardized weights**

Finally, we’ll compute the average neighbor GDPPC value for each polygon. These values are often referred to as **spatially lagged values**.

```{r}
GDPPC.lag <- lag.listw(rswm_q, hunan$GDPPC)
GDPPC.lag
```

Recall in the previous section, we retrieved the GDPPC of these five countries by using the code chunk below:

```{r}
nb1 <- wm_q[[1]]
nb1 <- hunan$GDPPC[nb1]
nb1
```

We can append the spatially lag GDPPC values onto hunan sf data frame by using the code chunk below:

```{r}
lag.list <- list(hunan$NAME_3, lag.listw(rswm_q, hunan$GDPPC))
lag.res <- as.data.frame(lag.list)
colnames(lag.res) <- c("NAME_3", "lag GDPPC")
hunan <- left_join(hunan,lag.res)
```

The following table shows the average neighboring income values (stored in the Inc.lag object) for each county.

```{r}
head(hunan)
```

Next, we will plot both the GDPPC and spatial lag GDPPC for comparison using the code chunk below:

```{r}
gdppc <- qtm(hunan, "GDPPC")
lag_gdppc <- qtm(hunan, "lag GDPPC")
tmap_arrange(gdppc, lag_gdppc, asp=1, ncol=2)
```

### **10.b. Spatial Lag as a Sum of Neighboring Values** 

Spatial lag can be computed as the **sum of neighboring values** by assigning **binary weights**.

Steps:

1.  Use the **neighbors list**.

2.  Apply a function (via **`lapply`**) to assign a weight of **1** to each neighbor.

3.  Use **`nb2listw()`** with `glist=` to explicitly assign these binary weights.

-   **`lapply`** applies the function to each element of the neighbors structure, facilitating manipulation.

```{r}
b_weights <- lapply(wm_q, function(x) 0*x + 1)
b_weights2 <- nb2listw(wm_q, 
                       glist = b_weights, 
                       style = "B")
b_weights2
```

With the proper weights assigned, we can use lag.listw to compute a lag variable from our weight and GDPPC.

```{r}
lag_sum <- list(hunan$NAME_3, lag.listw(b_weights2, hunan$GDPPC))
lag.res <- as.data.frame(lag_sum)
colnames(lag.res) <- c("NAME_3", "lag_sum GDPPC")
```

First, let us examine the result by using the code chunk below.

```{r}
lag_sum
```

Next, we will append the *lag_sum GDPPC* field into `hunan` sf data frame by using the code chunk below:

```{r}
hunan <- left_join(hunan, lag.res)
```

Now, We can plot both the *GDPPC* and *Spatial Lag Sum GDPPC* for comparison using the code chunk below:

```{r}
gdppc <- qtm(hunan, "GDPPC")
lag_sum_gdppc <- qtm(hunan, "lag_sum GDPPC")
tmap_arrange(gdppc, lag_sum_gdppc, asp=1, ncol=2)
```

### **10.c. Spatial window average**

The spatial window average uses row-standardized weights and includes the diagonal element. To do this in R, we need to go back to the neighbors structure and add the diagonal element before assigning weights.

To add the diagonal element to the neighbour list, we just need to use *include.self()* from **spdep**.

```{r}
wm_qs <- include.self(wm_q)
```

Notice that the Number of nonzero links, Percentage nonzero weights and Average number of links are 536, 6.921488 and 6.090909 respectively as compared to wm_q of 448, 5.785124 and 5.090909

Let us take a good look at the neighbour list of area \[1\] by using the code chunk below:

```{r}
wm_qs[[1]]
```

Notice that now \[1\] has six neighbours instead of five.

Now we obtain weights with *nb2listw()*

```{r}
wm_qs <- nb2listw(wm_qs)
wm_qs
```

Again, we use *nb2listw()* and *glist()* to explicitly assign weight values.

Lastly, we just need to create the lag variable from our weight structure and GDPPC variable.

```{r}
lag_w_avg_gpdpc <- lag.listw(wm_qs, 
                             hunan$GDPPC)
lag_w_avg_gpdpc
```

Next, we will convert the lag variable listw object into a data.frame by using *as.data.frame()*.

```{r}
lag.list.wm_qs <- list(hunan$NAME_3, lag.listw(wm_qs, hunan$GDPPC))
lag_wm_qs.res <- as.data.frame(lag.list.wm_qs)
colnames(lag_wm_qs.res) <- c("NAME_3", "lag_window_avg GDPPC")
```

Note: The third command line on the code chunk above renames the field names of *lag_wm_q1.res* object into *NAME_3* and *lag_window_avg GDPPC* respectively.

Next, the code chunk below will be used to append *lag_window_avg GDPPC* values onto *hunan* sf data.frame by using *left_join()* of **dplyr** package.

```{r}
hunan <- left_join(hunan, lag_wm_qs.res)
```

To compare the values of lag GDPPC and Spatial window average, `kable()` of Knitr package is used to prepare a table using the code chunk below.

```{r}
hunan %>%
  select("County", 
         "lag GDPPC", 
         "lag_window_avg GDPPC") %>%
  kable()
```

Lastly, *qtm()* of **tmap** package is used to plot the lag_gdppc and w_ave_gdppc maps next to each other for quick comparison.

```{r}
w_avg_gdppc <- qtm(hunan, "lag_window_avg GDPPC")
tmap_arrange(lag_gdppc, w_avg_gdppc, asp=1, ncol=2)
```

Note: For more effective comparison, it is advicible to use the core tmap mapping functions.

### **10.d. Spatial window sum**

The spatial window sum is the counter part of the window average, but without using row-standardized weights.

To add the diagonal element to the neighbour list, we just need to use *include.self()* from **spdep**.

```{r}
wm_qs <- include.self(wm_q)
wm_qs
```

Next, we will assign binary weights to the neighbour structure that includes the diagonal element.

```{r}
b_weights <- lapply(wm_qs, function(x) 0*x + 1)
b_weights[1]
```

Notice that now \[1\] has six neighbours instead of five.

Again, we use *nb2listw()* and *glist()* to explicitly assign weight values.

```{r}
b_weights2 <- nb2listw(wm_qs, 
                       glist = b_weights, 
                       style = "B")
b_weights2
```

With our new weight structure, we can compute the lag variable with *lag.listw()*.

```{r}
w_sum_gdppc <- list(hunan$NAME_3, lag.listw(b_weights2, hunan$GDPPC))
w_sum_gdppc
```

Next, we will convert the lag variable listw object into a data.frame by using *as.data.frame()*.

```{r}
w_sum_gdppc.res <- as.data.frame(w_sum_gdppc)
colnames(w_sum_gdppc.res) <- c("NAME_3", "w_sum GDPPC")
```

Note: The second command line on the code chunk above renames the field names of *w_sum_gdppc.res* object into *NAME_3* and *w_sum GDPPC* respectively.

Next, the code chunk below will be used to append *w_sum GDPPC* values onto *hunan* sf data.frame by using *left_join()* of **dplyr** package.

```{r}
hunan <- left_join(hunan, w_sum_gdppc.res)
```

To compare the values of lag GDPPC and Spatial window average, `kable()` of Knitr package is used to prepare a table using the code chunk below.

```{r}
hunan %>%
  select("County", "lag_sum GDPPC", "w_sum GDPPC") %>%
  kable()
```

Note: For more effective comparison, it is advicible to use the core tmap mapping functions.

## **10. References**

-   <https://cran.r-project.org/web/packages/spdep/vignettes/nb_sf.html>

git add -A
